---
title: "STAT 506 Problem Set 1"
format:
  html:
    code-fold: true
    embed-resources: true
---

# Problem 1: Exploratory Analysis of Soybean Cultivars

## (a) Data source and import

Data source: [Forty soybean cultivars from subsequent harvests, UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/913/forty+soybean+cultivars+from+subsequent+harvests).

The observations are in `data.csv`. I use the variable names in the UCI documentation; the accompanying `cultivars-description.ods` describes the cultivars but is not needed for these calculations.

```{r}
soybean <- read.csv(
  "data.csv",
  header = TRUE,
  stringsAsFactors = FALSE
)

names(soybean) <- c(
  "Season", "Cultivar", "Repetition", "PH", "IFP", "NLP",
  "NGP", "NGL", "NS", "MHG", "GY"
)

str(soybean)
head(soybean, 3)
```

## (b) Verify the dataset structure

```{r}
data.frame(
  observations = nrow(soybean),
  distinct_cultivars = length(unique(soybean$Cultivar)),
  distinct_seasons = length(unique(soybean$Season))
)
```

The data contain 320 observations from 40 cultivars across two seasons.

### Observations by season

```{r}
season_counts <- as.data.frame(table(soybean$Season))
names(season_counts) <- c("Season", "Observations")
knitr::kable(season_counts, caption = "Observations in each season")
```

Each season contributes 160 observations.

## (c) Questions about the soybean data

### 1. Correlation between plant height (PH) and grain yield (GY)

```{r}
cor(soybean$PH, soybean$GY)
```

The overall Pearson correlation is approximately 0.1233, a weak positive linear association between plant height and grain yield.

### 2. Correlation between PH and GY by season

```{r}
season_correlations <- sapply(
  split(soybean, soybean$Season),
  function(season_data) cor(season_data$PH, season_data$GY)
)
season_correlations
```

The correlation is approximately 0.2309 in Season 1 and -0.0227 in Season 2. Season 1 has the higher correlation.

### 3. Grain yield at the greatest plant height

```{r}
soybean[which.max(soybean$PH), c("Season", "Cultivar", "Repetition", "PH", "GY")]
```

The maximum recorded plant height was 94.8 cm; grain yield for that observation was approximately 3304.68.

### 4. Percentage with above-average thousand seed weight (MHG)

```{r}
mean_mhg <- mean(soybean$MHG)
above_mean_mhg <- sum(soybean$MHG > mean_mhg)

data.frame(
  overall_mean_mhg = mean_mhg,
  observations_above_mean = above_mean_mhg,
  percentage_above_mean = 100 * above_mean_mhg / nrow(soybean)
)
```

The mean MHG is 168.32 g. Of the 320 observations, 151 exceed that mean, or 47.2%.

### 5. Cultivar with the highest average grain yield

```{r}
mean_yield_by_cultivar <- aggregate(GY ~ Cultivar, data = soybean, FUN = mean)
mean_yield_by_cultivar[which.max(mean_yield_by_cultivar$GY), ]
```

Cultivar 98R30 CE has the highest average grain yield, approximately 3964.97.

## (d) Mean of every quantitative variable

Season and Repetition identify the study design; the table summarizes the eight measured quantitative variables.

```{r}
measure_vars <- c("PH", "IFP", "NLP", "NGP", "NGL", "NS", "MHG", "GY")

overall_means <- data.frame(
  Group = "Overall",
  t(colMeans(soybean[measure_vars])),
  check.names = FALSE
)
season_means <- aggregate(
  soybean[measure_vars],
  by = list(Group = paste("Season", soybean$Season)),
  FUN = mean
)

mean_table <- rbind(overall_means, season_means)
knitr::kable(
  mean_table,
  digits = 2,
  row.names = FALSE,
  caption = "Overall and season-specific means of quantitative variables"
)
```

Mean GY was 3428.45 in Season 1 and 3408.66 in Season 2. Mean MHG was 162.29 in Season 1 and 174.36 in Season 2.

## (e) Grain yield comparison between seasons

### Hypotheses and testing approach

For the Welch two-sample test, the null hypothesis is equal mean GY in the two seasons; the two-sided alternative is a difference. Welch's test allows unequal variances. As a sensitivity analysis, I average repetitions within each cultivar and season, then use a paired t-test on the 40 cultivar-level differences. Its null hypothesis is a mean within-cultivar difference of zero.

### R output

```{r}
gy_season_1 <- soybean$GY[soybean$Season == 1]
gy_season_2 <- soybean$GY[soybean$Season == 2]

data.frame(
  Season = c(1, 2),
  n = c(length(gy_season_1), length(gy_season_2)),
  mean_GY = c(mean(gy_season_1), mean(gy_season_2)),
  sd_GY = c(sd(gy_season_1), sd(gy_season_2))
)

welch_test <- t.test(gy_season_1, gy_season_2)

cultivar_season_means <- aggregate(
  GY ~ Cultivar + Season,
  data = soybean,
  FUN = mean
)
paired_means <- reshape(
  cultivar_season_means,
  idvar = "Cultivar",
  timevar = "Season",
  direction = "wide"
)
paired_test <- t.test(
  paired_means$GY.1,
  paired_means$GY.2,
  paired = TRUE
)

welch_test
paired_test
```

### Interpretation

Mean GY was 19.80 higher in Season 1. For the Season 1 minus Season 2 difference, the Welch test gave p = 0.7255 and a 95% confidence interval from -91.16 to 130.75. The cultivar-level paired test gave p = 0.8573. Neither test provides evidence of a season difference at the 5% level; these results do not establish equivalence.

# Problem 2: Adult Income Dataset

## (a) Data source and import

Data source: [Adult, UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/2/adult).

I combine `adult/adult.data` and `adult/adult.test` to analyze all 48,842 records. The test file begins with a non-data line, which is skipped during import.

```{r}
adult_names <- c(
  "age", "workclass", "fnlwgt", "education", "education-num",
  "marital-status", "occupation", "relationship", "race", "sex",
  "capital-gain", "capital-loss", "hours-per-week",
  "native-country", "income"
)

adult_train <- read.csv(
  file.path("adult", "adult.data"),
  header = FALSE,
  col.names = adult_names,
  check.names = FALSE,
  strip.white = TRUE,
  na.strings = "?",
  stringsAsFactors = FALSE
)
adult_test <- read.csv(
  file.path("adult", "adult.test"),
  header = FALSE,
  skip = 1,
  col.names = adult_names,
  check.names = FALSE,
  strip.white = TRUE,
  na.strings = "?",
  stringsAsFactors = FALSE
)

adult <- rbind(adult_train, adult_test)
data.frame(
  portion = c("Training", "Test", "Combined"),
  observations = c(nrow(adult_train), nrow(adult_test), nrow(adult))
)
str(adult)
```

## (b) Clean and simplify variable names

```{r}
names(adult) <- gsub("-", "_", names(adult), fixed = TRUE)
names(adult)[names(adult) == "fnlwgt"] <- "final_weight"
names(adult)
```

The names now follow snake_case; `fnlwgt` is also renamed `final_weight`.

## (c) Examine missing values

```{r}
missing_counts <- colSums(is.na(adult))
missing_table <- data.frame(
  variable = names(missing_counts),
  missing_observations = as.integer(missing_counts)
)
knitr::kable(missing_table, caption = "Missing observations by variable")
```

Workclass has 2,799 missing values, occupation has 2,809, and native country has 857. The other 12 variables have none.

## (d) Restrict to at least 10 hours worked per week

```{r}
n_before_hours <- nrow(adult)
adult_10plus <- adult[adult$hours_per_week >= 10, ]

data.frame(
  stage = c("Before hours filter", "At least 10 hours per week"),
  observations = c(n_before_hours, nrow(adult_10plus))
)
```

Restricting the sample to at least 10 hours of work per week reduces it from 48,842 to 48,142 observations.

## (e) Remove specified missing observations

```{r}
required_fields <- c("occupation", "workclass", "native_country")
adult_complete <- adult_10plus[
  complete.cases(adult_10plus[required_fields]),
]

data.frame(
  stage = c("After hours filter", "After missing-value filter"),
  observations = c(nrow(adult_10plus), nrow(adult_complete))
)
```

Excluding records with an unknown occupation, workclass, or native country produces a sample complete for these three variables. It leaves 44,777 observations, 3,365 fewer than after the hours filter. If missingness is systematic, this restriction may affect the sample's representativeness.

## (f) Apply age and education exclusions

```{r}
data.frame(
  exclusion = c("Age below 18", "Age above 90", "Preschool education"),
  observations = c(
    sum(adult_complete$age < 18),
    sum(adult_complete$age > 90),
    sum(adult_complete$education == "Preschool")
  )
)

adult_final <- adult_complete[
  adult_complete$age >= 18 &
    adult_complete$age <= 90 &
    adult_complete$education != "Preschool",
]
```

Of the remaining records, 454 have age below 18, none have age above 90, and 72 have Preschool education. Applying these exclusions removes 526 records.

## (g) Final sample size

```{r}
nrow(adult_final)
```

The final sample contains 44,251 observations.

# Problem 3: Happy Numbers

For a positive integer, the next value in its happy-number sequence is the sum of the squares of its digits. Repeating this process either reaches 1 or enters a cycle that does not include 1.

## (a) nextHappy()

### Function documentation and definition

```{r}
#' Calculate the next value in a happy-number sequence
#'
#' @param n One finite, positive whole number.
#' @return A positive integer: the sum of the squares of the digits of n.
nextHappy <- function(n) {
  if (length(n) != 1L || !is.numeric(n) || !is.finite(n) ||
      n <= 0 || n != floor(n)) {
    stop("n must be one positive integer.", call. = FALSE)
  }

  total <- 0L
  while (n > 0) {
    digit <- n %% 10
    total <- total + as.integer(digit^2)
    n <- n %/% 10
  }

  total
}
```

### Reproduce the supplied examples

```{r}
nextHappy(19)
nextHappy(82)
```

The examples return 82 and 68, as required.

## (b) happySequence()

### Function documentation and definition

```{r}
#' Follow a happy-number sequence until it reaches 1 or repeats
#'
#' @param n One finite, positive whole number.
#' @return A list with sequence (the visited values, starting at n) and
#'   isHappy (TRUE if the sequence reaches 1, FALSE if a value repeats).
happySequence <- function(n) {
  if (length(n) != 1L || !is.numeric(n) || !is.finite(n) ||
      n <= 0 || n != floor(n)) {
    stop("n must be one positive integer.", call. = FALSE)
  }

  sequence <- n
  while (tail(sequence, 1L) != 1) {
    next_value <- nextHappy(tail(sequence, 1L))
    if (next_value %in% sequence) {
      return(list(sequence = sequence, isHappy = FALSE))
    }
    sequence <- c(sequence, next_value)
  }

  list(sequence = sequence, isHappy = TRUE)
}
```

The function applies `nextHappy()` until the sequence reaches 1 or is about to repeat a value. It does not append the repeated value; this convention also determines sequence length in part (c).

### Reproduce the supplied examples

```{r}
happySequence(19)
happySequence(4)
```

Starting at 19 reaches 1 through 19, 82, 68, 100, 1, so `isHappy` is TRUE. Starting at 4 eventually returns to 4, so `isHappy` is FALSE.

## (c) Happy numbers among integers 1 through 1000

### Number of happy integers

```{r}
happy_results <- lapply(1:1000, happySequence)
is_happy <- vapply(
  happy_results,
  function(result) result$isHappy,
  logical(1)
)
sum(is_happy)
```

Of the integers from 1 through 1000, 143 are happy.

### Percentage of happy integers

```{r}
100 * sum(is_happy) / length(is_happy)
```

The percentage is 14.3%, using all 1,000 starting integers as the denominator.

### Longest happy sequence before reaching 1

```{r}
happy_starts <- which(is_happy)
happy_lengths <- vapply(
  happy_results[is_happy],
  function(result) length(result$sequence),
  integer(1)
)
longest_length <- max(happy_lengths)
longest_starts <- happy_starts[happy_lengths == longest_length]

data.frame(
  starting_value = longest_starts,
  sequence_length = longest_length
)
```

The longest sequence among happy numbers has seven values before reaching 1. There is no unique starting value: 21 integers tie, as shown in the table.

# Sources and acknowledgments

The [UCI soybean dataset](https://archive.ics.uci.edu/dataset/913/forty+soybean+cultivars+from+subsequent+harvests) and [UCI Adult dataset](https://archive.ics.uci.edu/dataset/2/adult) supplied the data and variable definitions. ChatGPT assisted with reviewing the code style and text.
